Transformation of forecasts for evaluating predictive performance in an epidemiological context

### 2023-01-12 main.tex

% Suggestion J:
% Forecast evaluation plays an essential role in the development cycle of pedictive epidemic models and can inform their use for public health decision making. Common scores to evaluate epidemiological forecasts are the Continuous Ranked Probability Score (CRPS) and the Weighted Interval Score (WIS), which are both measures of the absolute distance between the forecast distribution and the observation. They are commonly applied directly to predicted and observed incidence counts, but it can be questioned whether this yields the most meaningful results given the exponential nature of epidemic processes and the fact that observed values over space and time can span several orders of magnitude. In this paper we argue that applying the CRPS to incidence counts on a logarithmic rather than natural scale effectively mitigates these difficulties, yielding epidemiologically meaningful and easily interpretable results. We motivate the procedure threefold: firstly, it can be interpreted as a probabilistic version of a relative error. Secondly, it reflects how well models predicted the time-varying epidemic growth rate. And lastly, using arguments on variance-stabilizing transformations, it can be shown that under the assumption of a quadratic mean-variance relationship, the logarithmic transformation leads to expected CRPS values which are independent of the order of magnitude of the predicted quantity. This implies that different forecasting tasks will will have a similar impact on overall evaluation results. Applying the log-transformation to data from the European COVID-19 Forecast Hub, we find that it changes model rankings, regardless of stratification by forecast date, location or target types. Situations in which models missed beginning upward swings are more strongly emphasized, while penalising missed peaks are less severely penalized. Also, we show that the assumption of a quadratic mean-variance relationship is quite closely aligned with empirical patterns.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{output/figures/SIM-mean-state-size.png}
    \caption{Expected CRPS scores as a function of the mean and variance of the forecast quantity. We generated 10,000 samples for three distributions with differing relationships between mean and variance. We then computed expected CRPS values assuming an ideal forecaster with predictive distribution equal to the data-generating distribution. The variance of the negative binomial is given as $\sigma^2 = \mu + \mu^2 / \theta$, meaning that for large theta the negative binomial distribution is equal to the Poisson distribution. We used values of $\theta = 1$ (red), $\theta = \frac{1000}{\sqrt{\mu}}$ (green) and $\theta = 10^9$ (blue, approximately corresponding to a Poisson distribution). To make the scores for the different distributions comparable, scores were normalised to one, meaning that the mean score for every distribution (red, green, blue) is one. 
    A: Normalised expected CRPS for ideal forecasts with increasing means for three distribution with different relationships between mean and variance. B: A but with x axis on the log scale.}
    \label{fig:SIM-wis-state-size-mean}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width = 1\textwidth]{output/figures/illustration-effect-log-ranking-crps.png}
\caption{Illustration of the effect of the log-transformation of the ranking for a single forecast. Shown are CRPS (or WIS, respectively) values as a function of the observed value for two forecasters. Model A issues a negative binomial forecast with mean = 10 and $\theta = 1$ (resulting in a variance of 110), while Model B issues a Poisson distribution with mean = 10 (resulting in a variance of 10).}
\label{fig:illustration-ranking}
\end{figure}

@@

To check this, we ran the regressions
\begin{equation*}
    \text{CRPS}(F_{\log}, \log y) = \alpha_{\log} + \beta_{\log} \cdot \log{(\text{median}(F))},
\end{equation*}
where $F_{\log}$ is the predictive distribution for $\log(y)$, and
\begin{equation*}
    \text{CRPS}(F_{\sqrt{\ }}, \sqrt{y}) = \alpha_{\sqrt{\ }} + \beta_{\sqrt{\ }} \cdot \sqrt{\text{median}(F)},
\end{equation*} 
where $F_{\sqrt{\ }}$ is the predictive distribution on the square-root scale. As can be seen from Table \ref{tab:HUB-regression}, the results indeed indicate small targets lead to larger average CRPS when using the log transform ($\beta_{\log} < 0$), while the opposite is true for the square-root transform ($\beta_{\sqrt{\ }} > 0$). The results of the three regressions are also displayed in Figure \ref{fig:HUB-cors}. It appears that the log transformation is able to stabilise CRPS value in this empirical example.

@@

In this paper we proposed the use of transformations, with a particular focus on the natural logarithmic transformation, when evaluating forecasts in an epidemiological setting. These transformations can address issues that arise when evaluating epidemiological forecasts based on measures of absolute error and their probabilistic generalisations (i.e CRPS and WIS). We showed that scores obtained after log-transforming both forecasts and observations can be interpreted as a) a measure of relative prediction errors, as well as b) a score for a forecast of the exponential growth rate of the target quantity and c) as variance stabilising transformation in some settings.
When applying this approach to forecasts from the European COVID-19 Forecast Hub, we found overall scores on the log scale to be more equal across, time, location and target type (cases, deaths) than scores on the natural scale. Scores on the log scale were much less influenced by the overall incidence level in a country, and showed a slight tendency to be higher in locations with very low incidences. We found that model rankings changed noticeably. 
%, favouring models like epiforecasts-EpiNow2 which uses a log scale generative process and a negative binomial observation model, and disfavouring models like MUNI-ARIMA which uses a square root transform to stabilise variance. 
On the natural scale, missing the peak and overshooting was more severely penalised than missing the nadir and the following upswing in numbers. Both failure modes tended to be more equally penalised on the log scale (with undershooting receiving slightly higher penalties in our example).

Applying a log-transformation prior to the WIS means that forecasts are evaluated in terms of relative errors and errors on the exponential growth rate, rather than absolute errors. The most important strength of this approach is that the evaluation better accommodates the exponential nature of epidemiological process and the types of errors forecasters who accurately model those processes are expected to make. The log-transformation also helps avoid issues with scores being strongly influenced by the order of magnitude of the forecast quantity, which can be an issue when evaluating forecasts on the natural scale. 
A potential downside is that forecast evaluation is unreliable in situations where observed values are zero or very small. Including very small values in prediction intervals (see e.g. Figure \ref{fig:HUB-model-comparison-baseline}) can lead to excessive dispersion values on the log scale. 
Similarly, locations with lower incidences may get disproportionate weight (i.e. high scores) when evaluating forecasts on the log scale. \cite{bracherEvaluatingEpidemicForecasts2021} argue that the large weight given to forecasts for locations with high incidences is a desirable property, as it reflects performance on the targets we should care about most. On the other hand, scoring forecasts on the log scale may be less influenced by outliers and better reflect consistent performance across time, space, and forecast targets. It also gives higher weight to another type of situation one may care about, namely one in which numbers start to rise from a previously low level. 
If one is interested in scoring a forecast of the multiplicative, rather than the exponential growth rate, it may be better to use a different transformation, e.g. dividing every forecast and corresponding observed value by the previously last observed value when the forecast was made.

The log-transformation is only one of many transformations that may be useful and appropriate in an epidemiological context. One obvious option is to apply a population standardization to obtain incidence forecasts e.g., per 100,000 population If one is interested in multiplicative, rather than exponential growth rates, one could convert forecasts into forecasts for the multiplicative growth rate by dividing numbers by the last observed value. We suggested to use the natural logarithm as a variance-stabilising transformation (or alternatively the square-root transformation in the case of a Poisson distributed variable). Other VST like the Box-Cox \citep{boxAnalysisTransformations1964} are conceivable as well. Another promising transformation would be take differences between forecasts on the log scale, or alternatively
to divide each forecast by the forecast of the previous week (and analogously for observations), in order to obtain forecasts for week-to-week growth rates. One could then also ask forecasters to provide estimates of the weekly relative change applied to the latest data and subsequent forecast points directly. This would be akin to evaluating the shape of the predicted trajectory against the shape of the observed trajectory \citep[for a different approach to evaluate the shape of a forecast, see][]{srivastavaShapebasedEvaluationEpidemic2022}. This, unfortunately is not feasible in the context of forecasts stored as predictive quantiles, as the growth rate of the $\alpha$-quantile may be different from the $\alpha$-quantile of the growth-rate, but may be an interesting approach if predictive samples are available. What makes this approach of transforming forecasts before scoring particularly interesting is the possibility to construct composite scores as a weighted sum of scores based on different transformations. This would allow forecast consumers to assign explicit weights to different qualities of the forecasts they might care about.

In this work we focused on the CRPS and WIS, which are widely used in the evaluation of epidemic forecasts. We note that for the logarithmic score, which has also been used e.g., in some editions of the FluSight challenge \cite{reichCollaborativeMultiyearMultimodel2019}, the question of the right scale to evaluate forecasts does not arise. It is known that log score differences between different forecasters are invariant to monotonic transformations of the outcome variable (see e.g., \citealt{Diks2011}). This is clearly an advantage of the logarithmic score over the CRPS; however, the logarithmic score is known to have other severe downsides, e.g., its low robustness to occasional misguided forecasts; see \cite{bracherEvaluatingEpidemicForecasts2021} for a more detailed discussion.

Exploring transformations is a promising avenue for future work that could help bridge the gap between modellers and policy makers by providing scoring rules that better reflect what forecast consumers care about. Potentially, the variance stabilising time-series forecasting literature may be a useful source of transformations for various forecast settings. We have showed that the log-transformation can lead to significant changes in the relative rankings of models against each other, with potentially important implications for decision-makers who rely on the knowledge of past performance to make a decision about which forecasts should inform future decisions. While it is commonly accepted that multiple proper scoring rules should usually be considered when comparing forecasts, we think this should be supplemented by considering different transformations of the data to obtain a richer picture of model performance. More work needs to be done to better understand the effects of applying transformations more comprehensively and in different contexts, and how it might affect decision-making.

@@

\begin{table}
\centering
\begin{tabular}{ccrrrrrr}
\toprule
Horizon & Target & $\alpha$ & $\beta$ & $\alpha^*$ & $\beta^*$ & $\alpha^{**}$ & $\beta^{**}$\\
\midrule
all & all & -1.093 & 0.963 & 0.390 & 0.001 & -0.995 & 0.209\\
\addlinespace
all & Cases & 0.036 & 0.858 & 0.751 & -0.033 & 0.047 & 0.201\\
all & Deaths & -0.885 & 0.868 & 0.430 & -0.021 & -7.564 & 0.826\\
\addlinespace
1 & all & -1.402 & 0.923 & 0.318 & -0.014 & 0.318 & 0.088\\
2 & all & -1.221 & 0.967 & 0.364 & -0.003 & 0.111 & 0.164\\
3 & all & -1.001 & 0.984 & 0.408 & 0.008 & -1.283 & 0.255\\
4 & all & -0.757 & 0.986 & 0.467 & 0.016 & -1.041 & 0.312\\
\addlinespace
1 & Cases & -0.861 & 0.876 & 0.433 & -0.024 & 0.785 & 0.087\\
2 & Cases & -0.243 & 0.877 & 0.660 & -0.031 & 0.959 & 0.162\\
3 & Cases & 0.372 & 0.855 & 0.882 & -0.037 & 1.115 & 0.238\\
4 & Cases & 0.816 & 0.837 & 1.009 & -0.036 & 1.665 & 0.296\\
\addlinespace
1 & Deaths & -1.145 & 0.831 & 0.376 & -0.035 & 0.458 & 0.048\\
2 & Deaths & -0.980 & 0.866 & 0.415 & -0.027 & 0.445 & 0.084\\
3 & Deaths & -0.809 & 0.885 & 0.441 & -0.016 & -8.262 & 0.903\\
4 & Deaths & -0.603 & 0.891 & 0.488 & -0.007 & -7.834 & 0.898\\
\bottomrule
\end{tabular}
\caption{Coefficients of three regressions for the effect of the magnitude of the median forecast on expected scores. The first regression was 
$\text{CRPS}(F, y) = \exp(\alpha) \cdot \text{median}(F)^{\beta},$ where $F$ is the predictive distribution and $y$ the observed value. The second one was 
$\text{CRPS}(\log F, \log y) = \alpha^* + \beta^* \cdot \log{(\text{median}(F))}.$ The third one was $\text{CRPS}(\sqrt{F}, \sqrt{y}) = \alpha^{**} + \beta^{**} \cdot \sqrt{(\text{median}(F))}.$
}
\label{tab:HUB-regression}
\end{table}

>>>>

% Suggestion J:
% Forecast evaluation plays an essential role in the development cycle of predictive epidemic models and can inform their use for public health decision-making. Common scores to evaluate epidemiological forecasts are the Continuous Ranked Probability Score (CRPS) and the Weighted Interval Score (WIS), which are both measures of the absolute distance between the forecast distribution and the observation. They are commonly applied directly to predicted and observed incidence counts, but it can be questioned whether this yields the most meaningful results given the exponential nature of epidemic processes and the several orders of magnitude that observed values can span over space and time. In this paper, we argue that transforming counts before applying scores such as the CRPS or WIS can effectively mitigate these difficulties and yield epidemiologically meaningful and easily interpretable results. We motivate the procedure threefold using the CRPS on log-transformed counts as an example: firstly, it can be interpreted as a probabilistic version of a relative error. Secondly, it reflects how well models predicted the time-varying epidemic growth rate. And lastly, using arguments on variance-stabilizing transformations, it can be shown that under the assumption of a quadratic mean-variance relationship, the logarithmic transformation leads to expected CRPS values which are independent of the order of magnitude of the predicted quantity. This implies that different forecasting tasks will have a similar impact on overall evaluation results. Applying the log transformation to data and forecasts from the European COVID-19 Forecast Hub, we find that it changes model rankings regardless of stratification by forecast date, location or target types. Situations in which models missed beginning upward swings are more strongly emphasized while failing to predict a downturn following a peak is less severely penalized. We conclude that appropriate transformations, of which the natural logarithm is only one particularly attractive option, should be considered when assessing the performance of different models in the context of infectious disease incidence.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{output/figures/SIM-mean-state-size.png}
    \caption{Expected CRPS scores as a function of the mean and variance of the forecast quantity. We generated 10,000 samples for three distributions with differing relationships between mean and variance. We then computed expected CRPS values assuming an ideal forecaster with predictive distribution equal to the data-generating distribution. The first distribution (red) is a truncated normal distribution with constant variance (we chose a $\sigma = 1$ in order to only obtain positive samples). The second (green) is a geometric distribution (a negative binomial distribution with variance $\sigma^2 = \mu + \mu^2$. The third (blue) is a
    Poisson distribution with $\sigma^2 = \mu + \mu$. To make the scores for the different distributions comparable, scores were normalised to one, meaning that the mean score for every distribution (red, green, blue) is one. 
    A: Normalised expected CRPS for ideal forecasts with increasing means for three distribution with different relationships between mean and variance. B: A but with x axis on the log scale.}
    \label{fig:SIM-wis-state-size-mean}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width = 1\textwidth]{output/figures/illustration-effect-log-ranking-crps.png}
\caption{Illustration of the effect of the log-transformation of the ranking for a single forecast. Shown are CRPS (or WIS, respectively) values as a function of the observed value for two forecasters. Model A issues a geometric distribution with mean $\mu = 10$ and variance $\sigma^2 = \mu + \mu^2 = 110$), while Model B issues a Poisson distribution with mean and variance equal to 10. Zeroes in this illustrative example were handled by adding one before applying the natural logarithm.}
\label{fig:illustration-ranking}
\end{figure}

@@

To check this, we ran the regressions
\begin{equation*}
    \text{CRPS}(F_{\log}, \log y) = \alpha_{\log} + \beta_{\log} \cdot \log{(\text{median}(F))},
\end{equation*}
where $F_{\log}$ is the predictive distribution for $\log(y)$, and
\begin{equation*}
    \text{CRPS}(F_{\sqrt{\ }}, \sqrt{y}) = \alpha_{\sqrt{\ }} + \beta_{\sqrt{\ }} \cdot \sqrt{\text{median}(F)},
\end{equation*} 
where $F_{\sqrt{\ }}$ is the predictive distribution on the square-root scale. As can be seen from Table \ref{tab:HUB-regression}, the results indeed indicate small targets lead to larger average CRPS when using the log transform ($\beta_{\log} < 0$), while the opposite is true for the square-root transform ($\beta_{\sqrt{\ }} > 0$). The results of the three regressions are also displayed in Figure \ref{fig:HUB-cors}. It appears that the log transformation is most suitable to stabilise the CRPS values in this empirical example.

@@

In this paper, we proposed the use of transformations, with a particular focus on the natural logarithmic transformation, when evaluating forecasts in an epidemiological setting. These transformations can address issues that arise when evaluating epidemiological forecasts based on measures of absolute error and their probabilistic generalisations (i.e CRPS and WIS). We showed that scores obtained after log-transforming both forecasts and observations can be interpreted as a) a measure of relative prediction errors, as well as b) a score for a forecast of the exponential growth rate of the target quantity and c) as variance stabilising transformation in some settings.
When applying this approach to forecasts from the European COVID-19 Forecast Hub, we found overall scores on the log scale to be more equal across, time, location and target type (cases, deaths) than scores on the natural scale. Scores on the log scale were much less influenced by the overall incidence level in a country and showed a slight tendency to be higher in locations with very low incidences. We found that model rankings changed noticeably. 
%, favouring models like epiforecasts-EpiNow2 which uses a log scale generative process and a negative binomial observation model, and disfavouring models like MUNI-ARIMA which uses a square root transform to stabilise variance. 
On the natural scale, missing the peak and overshooting was more severely penalised than missing the nadir and the following upswing in numbers. Both failure modes tended to be more equally penalised on the log scale (with undershooting receiving slightly higher penalties in our example).

Applying a log-transformation prior to the WIS means that forecasts are evaluated in terms of relative errors and errors on the exponential growth rate, rather than absolute errors. The most important strength of this approach is that the evaluation better accommodates the exponential nature of the epidemiological process and the types of errors forecasters who accurately model those processes are expected to make. The log-transformation also helps avoid issues with scores being strongly influenced by the order of magnitude of the forecast quantity, which can be an issue when evaluating forecasts on the natural scale. 
A potential downside is that forecast evaluation is unreliable in situations where observed values are zero or very small. Including very small values in prediction intervals (see e.g. Figure \ref{fig:HUB-model-comparison-baseline}) can lead to excessive dispersion values on the log scale. 
Similarly, locations with lower incidences may get disproportionate weight (i.e. high scores) when evaluating forecasts on the log scale. \cite{bracherEvaluatingEpidemicForecasts2021} argue that the large weight given to forecasts for locations with high incidences is a desirable property, as it reflects performance on the targets we should care about most. On the other hand, scoring forecasts on the log scale may be less influenced by outliers and better reflect consistent performance across time, space, and forecast targets. It also gives higher weight to another type of situation one may care about, namely one in which numbers start to rise from a previously low level. 
If one is interested in scoring a forecast of the multiplicative, rather than the exponential growth rate, it may be better to use a different transformation, e.g. dividing every forecast and corresponding observed value by the previously last observed value when the forecast was made.

The log-transformation is only one of many transformations that may be useful and appropriate in an epidemiological context. One obvious option is to apply a population standardization to obtain incidence forecasts e.g., per 100,000 population If one is interested in multiplicative, rather than exponential growth rates, one could convert forecasts into forecasts for the multiplicative growth rate by dividing numbers by the last observed value. We suggested using the natural logarithm as a variance-stabilising transformation (VST) or alternatively the square-root transformation in the case of a Poisson distributed variable. Other VST like the Box-Cox \citep{boxAnalysisTransformations1964} are conceivable as well. Another promising transformation would be to take differences between forecasts on the log scale, or alternatively to divide each forecast by the forecast of the previous week (and analogously for observations), in order to obtain forecasts for week-to-week growth rates. One could then also ask forecasters to provide estimates of the weekly relative change applied to the latest data and subsequent forecast points directly. This would be akin to evaluating the shape of the predicted trajectory against the shape of the observed trajectory \citep[for a different approach to evaluating the shape of a forecast, see][]{srivastavaShapebasedEvaluationEpidemic2022}. This, unfortunately, is not feasible in the context of forecasts stored as predictive quantiles, as the growth rate of the $\alpha$-quantile may be different from the $\alpha$-quantile of the growth-rate, but may be an interesting approach if predictive samples are available as it enables the possibility of constructing composite scores as a weighted sum of scores based on different transformations. This would allow forecast consumers to assign explicit weights to different qualities of the forecasts they might care about.

In this work, we focused on the CRPS and WIS, which are widely used in the evaluation of epidemic forecasts. We note that for the logarithmic score, which has also been used e.g., in some editions of the FluSight challenge \cite{reichCollaborativeMultiyearMultimodel2019}, the question of the right scale to evaluate forecasts does not arise. It is known that log score differences between different forecasters are invariant to monotonic transformations of the outcome variable (see e.g., \citealt{Diks2011}). This is clearly an advantage of the logarithmic score over the CRPS; however, the logarithmic score is known to have other severe downsides, e.g., its low robustness to sporadically misguided forecasts; see \cite{bracherEvaluatingEpidemicForecasts2021} for a more detailed discussion.

Exploring transformations is a promising avenue for future work that could help bridge the gap between modellers and policymakers by providing scoring rules that better reflect what forecast consumers care about. Potentially, the variance stabilising time-series forecasting literature may be a useful source of transformations for various forecast settings. We have shown that the natural logarithm transformation can lead to significant changes in the relative rankings of models against each other, with potentially important implications for decision-makers who rely on the knowledge of past performance to make a judgement about which forecasts should inform future decisions. While it is commonly accepted that multiple proper scoring rules should usually be considered when comparing forecasts, we think this should be supplemented by considering different transformations of the data to obtain a richer picture of model performance. More work needs to be done to better understand the effects of applying transformations in different contexts, and how they may impact decision-making.

@@

\begin{table}
\centering
\begin{tabular}{ccrrrrrr}
\toprule
Horizon & Target & $\alpha$ & $\beta$ & $\alpha_{\log}$ & $\beta_{\log}$ & $\alpha_{\sqrt{\ }}$ & $\beta_{\sqrt{\ }}$\\
\midrule
all & all & -1.093 & 0.963 & 0.390 & 0.001 & -0.995 & 0.209\\
\addlinespace
all & Cases & 0.036 & 0.858 & 0.751 & -0.033 & 0.047 & 0.201\\
all & Deaths & -0.885 & 0.868 & 0.430 & -0.021 & -7.564 & 0.826\\
\addlinespace
1 & all & -1.402 & 0.923 & 0.318 & -0.014 & 0.318 & 0.088\\
2 & all & -1.221 & 0.967 & 0.364 & -0.003 & 0.111 & 0.164\\
3 & all & -1.001 & 0.984 & 0.408 & 0.008 & -1.283 & 0.255\\
4 & all & -0.757 & 0.986 & 0.467 & 0.016 & -1.041 & 0.312\\
\addlinespace
1 & Cases & -0.861 & 0.876 & 0.433 & -0.024 & 0.785 & 0.087\\
2 & Cases & -0.243 & 0.877 & 0.660 & -0.031 & 0.959 & 0.162\\
3 & Cases & 0.372 & 0.855 & 0.882 & -0.037 & 1.115 & 0.238\\
4 & Cases & 0.816 & 0.837 & 1.009 & -0.036 & 1.665 & 0.296\\
\addlinespace
1 & Deaths & -1.145 & 0.831 & 0.376 & -0.035 & 0.458 & 0.048\\
2 & Deaths & -0.980 & 0.866 & 0.415 & -0.027 & 0.445 & 0.084\\
3 & Deaths & -0.809 & 0.885 & 0.441 & -0.016 & -8.262 & 0.903\\
4 & Deaths & -0.603 & 0.891 & 0.488 & -0.007 & -7.834 & 0.898\\
\bottomrule
\end{tabular}
\caption{Coefficients of three regressions for the effect of the magnitude of the median forecast on expected scores. The first regression was 
$\text{CRPS}(F, y) = \exp(\alpha) \cdot \text{median}(F)^{\beta},$ where $F$ is the predictive distribution and $y$ the observed value. The second one was 
$\text{CRPS}(\log F, \log y) = \alpha^* + \beta^* \cdot \log{(\text{median}(F))}.$ The third one was $\text{CRPS}(\sqrt{F}, \sqrt{y}) = \alpha^{**} + \beta^{**} \cdot \sqrt{(\text{median}(F))}.$
}
\label{tab:HUB-regression}
\end{table}
